---
title: "Project 1"
author: "Ryan Fischbach"
date: "2/20/2021"
output: pdf_document
---

```{r setup, include=TRUE}
knitr::opts_chunk$set(echo = FALSE)
```

```{r}
library(ggplot2)
```

```{r}
#import data
starwars <- read.csv("~/Google Drive (fiscrt17@wfu.edu)/Fourth Year/STA363/Project 1/starwars.csv")

#subset data to only have columns of interest (mass, height, species)
subsetStarWars <- subset(starwars, select = c("mass", "height", "species"))
```

```{r}
#remove na values from subsetted dataframe
subsetStarWars = na.omit(subsetStarWars)
```

```{r out.width = "75%"}
#plot height of character on mass
ggplot(data = subsetStarWars, aes(x = mass, y = height)) + geom_point(color = "blue") + labs(x =" Mass of Character (kg)", y = "Height of Character (cm)", title = "Height of Character on Mass") + stat_smooth(method = "lm", formula = y~x, se = FALSE)
```

```{r out.width = "75%"}
#Create a new column for easier labeling
subsetStarWars$human <- subsetStarWars$species
subsetStarWars$human[subsetStarWars$species!='Human'] <- "Not Human"

#Graph relationship between height and mass with humans and not humans
ggplot(data = subsetStarWars, mapping = aes(x = subsetStarWars$mass, y = subsetStarWars$height, color = human)) + geom_point() + labs(x =" Mass of Character (kg)", y = "Height of Character (cm)", title = "Height of Character on Mass") + facet_wrap( ~ human, ncol = 2) + stat_smooth(method = "lm", formula = y~x, se = FALSE)

```

```{r}
#box plot to show outlier
#ggplot(subsetStarWars,  aes(  x = mass)  )  + geom_boxplot( fill = 'blue' )

#remove Hutt observation
subsetStarWars <- subsetStarWars[subsetStarWars$species != "Hutt",]
```

```{r}
#Create a new column for easier labeling
subsetStarWars$human <- subsetStarWars$species
subsetStarWars$human[subsetStarWars$species!='Human'] <- "Not Human"

#Graph relationship between height and mass with humans and not humans
ggplot(data = subsetStarWars, mapping = aes(x = subsetStarWars$mass, y = subsetStarWars$height, color = human)) + geom_point() + labs(x =" Mass of Character (kg)", y = "Height of Character (cm)", title = "Height of Character on Mass") + facet_wrap( ~ human, ncol = 2) + stat_smooth(method = "lm", formula = y~x, se = FALSE)

```
```


```{r}
#Count number of observations
n <- nrow(subsetStarWars)

#Prepare data for different model by creating binary dummy variable
subsetStarWars$human <- NA
subsetStarWars$human[subsetStarWars$species!='Human'] <- 0
subsetStarWars$human[subsetStarWars$species=='Human'] <- 1

#Set a Random Seed
set.seed(223)

#Choose training set indices using 80:20 split
rowsTraining <- sample(1:n, n*.8, replace=FALSE)

#Create training and test sets
VAtraining <- subsetStarWars[rowsTraining, ]
VAtest <- subsetStarWars[-rowsTraining, ]

#Fit model 1
combinedModel <- lm(height ~ mass, data=VAtraining)

#Fit model 2
differentModel <- lm(height ~ mass + human, data=VAtraining)

#calculate predictions
combined_predsVA <- predict(combinedModel, newdata=VAtest)
different_predsVA <- predict(differentModel, newdata=VAtest)

#Calculate rmse for the two models
rmse_m1VA = sqrt(mean((VAtest$height - combined_predsVA)^2))
rmse_m2VA = sqrt(mean((VAtest$height - different_predsVA)^2))

```


```{r}
#Choose k
k <- 10

#Define n
n <- nrow(subsetStarWars)

#Prepare data for different model by creating binary dummy variable
subsetStarWars$human <- NA
subsetStarWars$human[subsetStarWars$species!='Human'] <- 0
subsetStarWars$human[subsetStarWars$species=='Human'] <- 1

#Create storage
residualsKFOLD1 <- matrix(NA, nrow = n, ncol=1)
residualsKFOLD2 <- matrix(NA, nrow = n, ncol=1)

#Create folds
set.seed(123)

folds <- sample(rep(1:k, 40), n, replace=FALSE)

for(i in 1:k){
  #Find the rows in fold i
  infold <- which(folds==i)
  
  #Create training set
  CVTrain <- subsetStarWars[-infold,]
  
  #Create test set
  CVTest <- subsetStarWars[infold,]
  
  #Train the model(s) of interest.
  lmCV1 <- lm(height ~ mass, data=CVTrain)
  
  #Predict on CV Test data
  predCV1 <- predict(lmCV1, newdata=CVTest)
  
  #Compute the MSRE for each fold and store
  residualsKFOLD1[infold] <- sqrt(mean(CVTest$height - predCV1)^2)

  #Fit model 2 (separate variables human and nonhuman)
  CVdifferent_model <- lm(height ~ mass + human, data=CVTrain)
  
  #Compute predictions for human and nonhumans and combine
  CVdifferent_predictions <- predict(CVdifferent_model, newdata=CVTest)

  residualsKFOLD2[infold] <- sqrt(mean(CVTest$height - CVdifferent_predictions)^2)

}
  #Compute the average MSRE across all folds
  msre1 = mean(residualsKFOLD1)
  msre2 = mean(residualsKFOLD2)
```

```{r}
#train final model
finalModel = lm(height ~ mass + human, data=subsetStarWars)

#produce parameters of final model
#summary(finalModel)

#produce confidence interval of all slopes
#confint(finalModel)
```

